Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-21 paper-amia-summit-2019/manuscript.tex
edited by April and Alex up to page 5

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{soul}
\usepackage{multirow}
\renewcommand*{\thefootnote}{\alph{footnote}}

@@

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by the traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into fragments that correspond to MI behaviors. This study formulates this task as a classification problem and proposes word embeddings, punctuation and part-of-speech features to address it. We experimented both with traditional machine learning method, conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN). Results indicate that CRNN outperformed CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for detecting new segment.}

\section*{Introduction}
The emergence of e-Health technologies opened up new ways to deliver a variety of behavioral interventions to any demographic group of patients in any geographical location. Motivational interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change,\cite{miller2012motivational,miller2009toward} is one type of these interventions. MI sessions are generally aimed at eliciting ``change talk'', or statements of intrinsic motivation about patients' own desire, ability, reasons, need for and commitment to behavior change, which have been established by previous research\cite{apodaca2009mechanisms} as a reliable mediator of health behavior change. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods.

Qualitative analysis of motivational interviews has been traditionally performed manually by MI researchers, which is a tedious and resource-intensive process involving several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis.\cite{wang2012baselines} In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and predicting the outcome \cite{hasan2018predicting} of in-person MI sessions. Experimental data utilized in these studies, however, were transcribed auto recordings of in-person MI sessions with a counselor, which have a clear discourse structure and were clearly segmented into the utterances by a counselor, a patient or a caregiver.

In this study, we focus on the analysis of e-Coaching sessions, behavioral interventions that are delivered via email and are grounded in the principles of MI. Specifically, the e-Coaches involved in this study used emails to communicate motivation-enhancing messages that encourage healthy eating among GenY adolescents. E-coaching data is comprised of email responses, which are free-text documents, unlike more traditional dyadic clinical interviews that are naturally segmented into utterances due to their conversational nature.

The unstructured nature of e-Coaching exchanges poses a unique set of challenges for their qualitative analysis. Usually, email segments are automatically coded with the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE), \cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI session. A significant barrier to qualitative analysis of e-Coaching exchanges is their segmentation into textual fragments that correspond to a distinct e-Coach and patient communication behaviors. Automating this task is a unique and challenging problem due to the following reasons:

\begin{enumerate}
\item Emails are unstructured text containing informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several previous patient statements in one email. However, in a traditional MI interview session involving a dialogue between a provider and a patient, a provider utterance is a response to an immediately preceding inquiry or statement by a patient. 
\item Discourse segments in e-Coaching emails do not necessarily correspond to sentences or paragraphs. One sentence can be divided into fragments corresponding to multiple MI behaviors. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/segment-example.png}
    \caption{\textbf{Example of e-Coaching emails segmented into fragments that correspond to MI behaviors of an e-Coach and a patient}}
    \label{fig:text-segment}
\end{figure}

Figure~\ref{fig:text-segment} illustrates a segmentation of an e-Coaching email exchange, in which the first sentence is segmented into 2 MI behavior fragments, while the fourth and fifth segments correspond to one and three sentences, respectively. Segmentation of e-Coaching email exchanges between patients and e-Coaches corresponds to a special case of discourse analysis \cite{webber2012discourse} aimed at better understanding the effective communication strategies specific to this type of behavioral interventions and revealing the unique socio-psychological characteristics of patients.

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of qualitative analysis of this type of clinical communications. Specifically, for this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a recent \hl{email-based} public health intervention that relies on personalized e-Coaching to encourage increased fruit and vegetable intake among young adults, aged 21--30. The goal of MENU GenY was to develop a better coding scheme for e-Coaching communications aimed at improving GenY eating habits. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail based behavioral interventions}.

Specifically, we introduce and evaluate the effectiveness of word embedding or lexical, punctuation and part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning architectures, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

\section*{Relevant work}
Previous relevant work in the biomedical domain primarily focused on segmentation of text in electronic health records (EHR) into sections and headers \cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection.\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence} In particular, Maximum Entropy models \cite{tepper2012statistical} and Support Vector Machine (SVM) along with word-vector cosine similarity metric and several heuristics \cite{apostolova2009automatic} have been applied to segment or classify clinical documents in EHR into pre-defined sections, such as general patient information, medical history, procedures, findings, etc. Denny et al. \cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Naive Bayes classifier to identify sections and headers in EHR. Segmentation of e-Coaching emails, however, is different from segmentation of other clinical documents in that the focus is on clinical conversation.

SVM in conjunction with prosodic and part of speech features \cite{kreuzthaler2015detection} and Recurrent Convolutional Neural Networks \cite{griffis2016quantitative} have also been utilized for \textit{sentence boundary detection} in general text. Liu et al. \cite{liu2005using} demonstrated that a linear-chain CRF outperforms Hidden Markov and Maximum Entropy models for this task.

Segmentation of e-Coaching emails is different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers. The proposed methods will automate the process of segmenting clinical exchanges into groups of MI behaviors, which will significantly reduce the resources and time required to perform this task manually. Furthermore, these methods can be integrated with auto coding methods \cite{hasan2016study,kotov2015interpretable} to create a software pipeline for automated analysis of clinical conversation.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include \hl{330 and 281 emails from ecoaches and patients, respectively }(Table~\ref{tab:datastat}). \hl{Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and annotated using 115 distinct MYSCOPE}~\cite{carcone2013provider} \hl{behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances.} In this study, we experimented with traditional machine learning methods, such as Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing instances were created based on a sliding window of $2n$ words or punctuation marks over each position in a given input sequence, such that each instance consists of a pair of the next $n$ words or punctuation marks and prior $n$ words or punctuation marks, including the current word or punctuation mark. Each sample is classified into either a ``new segment'' or a ``same segment'' based on whether there should be a segment break after the current word or not. In the case of CRF, BRNN and CRNN models, an email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output of a model. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (an example is given at $2^{nd}$ row in Table~\ref{tab:datastat}). \\

\begin{table}[ht]
\centering
\caption{\textbf{\hl{Summary of the experimental dataset}}}
\label{tab:datastat}
 \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Sessions}} & \multirow{2}{*}{\textbf{Instances}} & \multicolumn{2}{|c|}{\textbf{Class Segments}} & \multicolumn{2}{|c|}{\textbf{Tokens}} & \multicolumn{2}{|c|}{\textbf{Emails}} & \multicolumn{2}{|c|}{\textbf{Annotation}} \\\cline{3-10}
   &  & \textbf{new}  & \textbf{same} & \textbf{words} & \textbf{punctuations}  & \textbf{patient} & \textbf{provider} & \textbf{method}  & \textbf{codes} \\ \hline    
 49 & 102,917 & 3,138 & 99,779 & 95,777 & 7,140 & 281 & 330 & MYSCOPE & 115 \\ \hline
 \multicolumn{10}{|c|}{segmentation example: Hi[0] XXX[0] ,[0] its[0] good[0] to[0] hear[0] from[0] you[0] .[1] it[0] sounds[0] like[0]...} \\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic units of semantics. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embedding is one such representation, when each word is associated with a real-valued vector in a high dimensional vector space. Distributed representations of words have been shown to capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized embeddings estimated on Google News corpus consisting of 1.6 billion words pre-trained using word2vec word embedding method.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} When words or punctuation marks are not found in the pre-trained word vectors, we utilized word embeddings trained with our e-Coaching email corpus, which we refer to as corpus-based embeddings. CRF utilized lexical features, which correspond to words in e-Coaching emails.

@@

We experimented with 4 different classifiers, including one traditional machine learning method, CRF and three deep learning models. Since deep learning architectures provide a flexible mechanism for constructing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Field}: CRF has been widely used in various NLP tasks, such as part-of-speech tagging.\cite{lafferty2001conditional, hirohata2008identifying} Unlike a maximum entropy Markov model, which uses per-state exponential models for conditional probability of the next state given a current state, CRF model directly estimates conditional distribution of the entire output sequence, given the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ for output and input sequences, $y$ and $x$:

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags.

\textbf{Multi-Layer Perceptron}: MLP is a feed-forward artificial neural network, which maps an input onto one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multilayer perceptron with a single hidden layer. In MLP networks, there is no cycle or loop and information moves forward only, from the input nodes through the hidden nodes and to the output nodes. This study utilized MLP with a nonlinear activation function (rectified linear unit) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we utilized dropout in fully connected layers by randomly hiding neurons during training. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

@@

\section*{Conclusion}
Segmentation of e-Coaching emails is an integral part of developing and analyzing e-Coaching behavioral interventions. Although several studies have focused on segmentation problem in biomedical context, they are limited to segmenting clinical text in EHR into sections and sentences, with none of them considering segmentation of text into groups of MI behaviors in the context of MI session discourse analysis. By comparing the performance of machine learning methods for the task of segmentation of e-Coaching emails, we found out that CRNN provides the best performance in terms of all performance metrics. Manual segmentation of e-Coaching data is a very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify textual segments corresponding to MI behaviors in unstructured clinical dialog, which can then be annotated with a classification model in a pipeline setting. Automated segmentation and analysis of e-Coaching emails can significantly decrease the time to identify effective communication strategies in e-mail based motivational interviewing.

>>>>

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{soul}
\usepackage{multirow}
\usepackage{url}
\renewcommand*{\thefootnote}{\alph{footnote}}

@@

\noindent{\bf Abstract}
\textit{Communication science approaches to develop effective behavior interventions, such as motivational interviewing (MI), are limited by traditional qualitative coding of communication exchanges, a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated qualitative coding of e-Coaching communication exchanges is segmentation of emails into fragments that correspond to MI behaviors. This study frames email segmentation task as a classification problem and utilizes word embeddings in conjunction with punctuation and part-of-speech features to address it. We experimentally evaluated conditional random fields (CRF), a traditional machine learning method, as well as deep learning methods, such as multilayer perceptron (MLP), bidirectional recurrent neural network (BRNN) and convolutional recurrent neural network (CRNN). Our results indicate that CRNN outperform CRF, MLP and BRNN for the task of email segmentation achieving 0.989 macro F1-score overall and 0.825 macro F1-score for new segment detection.}

\section*{Introduction}
The emergence of e-Health technologies has greatly expanded the reach of behavioral interventions. One such intervention is email-delivered Motivational Interviewing (MI). MI is an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009toward}. MI is linked to behavior change through the elicitation of patient ``change talk'', or statements of intrinsic motivation expressing patients’ own desires, abilities, reasons, need for, and commitment to behavior change\cite{apodaca2009mechanisms}. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods. The specific provider behaviors responsible for the elicitation of change talk, however, are less clear and may vary by treatment context. Thus, a current focus of MI research is understanding which provider behaviors in which contexts lead to patient change talk. To date, no study has examined this process in email-delivered MI.

The study of patient-provider communication in MI relies on the behavioral coding of intervention transcripts, an iterative, resource-intensive, and cognitively demanding process of qualitative analysis involving several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer an opportunity to dramatically accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis.\cite{wang2012baselines} In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and predicting the outcome \cite{hasan2018predicting} of in-person MI sessions. Experimental data utilized in these studies were transcribed auto recordings of in-person MI sessions with a counselor, which were segmented into counselor and client utterances during the transcription process.

In this study, we focus on the analysis of email-delivered MI, or e-Coaching, to promote healthy eating among young adults. The e-Coaching dataset is composed of email correspondence between a MI counselor and the young adult patient. Unlike transcribed in-person discourse, email correspondence is not clearly segmented into codable speech acts (i.e., utterances). Thus, the unstructured nature of e-Coaching exchanges poses a unique set of analytic challenges. Segmentation of e-Coaching exchanges into textual fragments that correspond to distinct e-Coach and patient communication behaviors is a significant barrier to qualitative analysis of this type of clinical conversation. Automating this task is a unique and challenging problem due to the following reasons:

\begin{enumerate}
\item Emails are unstructured text containing informal information exchange in a non-traditional format. For example, an e-Coach usually responds to several previous patient statements in one email. In contrast, in a traditional, in-person MI session, an utterance is assumed to be a response to the immediately preceding utterance.
\item Discourse segments in e-Coaching do not have a clear breakpoint, such as the end of a sentence or paragraph. One sentence might be divided into fragments corresponding to multiple MI behaviors. On the other hand, an MI behavior may comprise several sentences.
\end{enumerate}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/segment-example.png}
    \caption{\textbf{Example of an e-Coaching exchange segmented into fragments corresponding to MI behaviors of an e-Coach and a patient}}
    \label{fig:text-segment}
\end{figure}

Figure~\ref{fig:text-segment} illustrates a segmentation of an e-Coaching exchange, in which the first sentence is segmented into 2 MI behavior fragments, while the fourth and fifth segments correspond to one and three sentences, respectively. Segmentation of e-Coaching exchanges constitutes a special case of clinical discourse analysis \cite{webber2012discourse} aimed at better understanding the effective communication strategies specific to this type of behavioral interventions.

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual patient and provider behaviors.  For this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a web-delivered public health intervention with email-based coaching to encourage increased fruit and vegetable intake among young adults, aged 21–-30. A secondary goal of the MENU GenY project was to identify the specific communication strategies used by e-Coaches to elicit change talk for healthier eating among young adult patients. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail-based behavioral interventions}.

Specifically, we introduce and evaluate the effectiveness of distributed word representations (i.e. word embeddings) as well as lexical, punctuation and part-of-speech (POS) features in conjunction with both traditional supervised machine learning methods, such as linear-chain Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning architectures, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

\section*{Related work}

Prior work on textual segmentation in the biomedical domain primarily focused on sentence boundary detection \cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence} and segmentation of clinical documents in patients' electronic health records (EHR) into sections and headers. \cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} In particular, Maximum Entropy models \cite{tepper2012statistical} and Support Vector Machine (SVM) along with word-vector cosine similarity metric and several heuristics \cite{apostolova2009automatic} have been applied to identify specific sections in EHR, such as general patient information, medical history, procedures, findings, etc. Denny et al. \cite{denny2009evaluation} proposed SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Na\"{i}ve Bayes classifier to identify sections and headers in EHR. Segmentation of e-Coaching emails, however, is different from segmentation of other clinical documents, since the focus is on dialog acts in clinical conversation.

SVM in conjunction with prosodic and part-of-speech features \cite{kreuzthaler2015detection} and Recurrent Convolutional Neural Networks \cite{griffis2016quantitative} have also been utilized for \textit{sentence boundary detection} in general text. Liu et al. \cite{liu2005using} demonstrated that a linear-chain CRF outperforms Hidden Markov and Maximum Entropy models for this task.

Segmentation of e-Coaching emails is also different from traditional shallow discourse analysis \cite{galley2003discourse}, which besides identification of speech acts, also aims to determine the types of transitions between speech acts and labels speech acts with speakers who performed them in a multi-speaker conversation. The proposed methods will automate the process of segmenting clinical exchanges into MI behaviors, which will significantly reduce the time and resources required to perform such segmentation manually. Furthermore, these methods can be integrated with auto coding methods \cite{hasan2016study,kotov2015interpretable} to create a software pipeline for fully automated analysis of email-delivered behavioral interventions.

\section*{Materials and methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include 330 and 281 emails by e-Coaches and patients, respectively. Various statistics of experimental dataset are provided in Table~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and and annotated using the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE),~\cite{carcone2013provider} a qualitative coding scheme to characterize patient-provider communication during MI sessions with 115 distinct behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a class label ``new segment'' or ``same segment'' to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing samples were created based on a sliding window of $2n$ words or punctuation marks over each position (which could be a word or a punctuation mark) in a given input sequence, such that each instance consists of the $n$ words or punctuation marks after this position and $n$ words or punctuation marks prior to this position, including the position itself. Each sample is assigned a ``new segment'' or ``same segment'' label based on whether there should be a segment break after the current position or not. In the case of CRF, BRNN and CRNN models, an e-Coaching email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1 (\hl{an example of a sequence with assigned labels is provided in the $2^{nd}$ row of Table}~\ref{tab:datastat}). \\

\begin{table}[ht]
\centering
\caption{\textbf{Summary of statistics of experimental dataset.}}
\label{tab:datastat}
 \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Sessions}} & \multirow{2}{*}{\textbf{Instances}} & \multicolumn{2}{|c|}{\textbf{Class Segments}} & \multicolumn{2}{|c|}{\textbf{Tokens}} & \multicolumn{2}{|c|}{\textbf{Emails}} & \multicolumn{2}{|c|}{\textbf{Annotation}} \\\cline{3-10}
   &  & \textbf{new}  & \textbf{same} & \textbf{words} & \textbf{punctuations}  & \textbf{patient} & \textbf{provider} & \textbf{method}  & \textbf{codes} \\ \hline    
 49 & 102,917 & 3,138 & 99,779 & 95,777 & 7,140 & 281 & 330 & MYSCOPE & 115 \\ \hline
 \multicolumn{10}{|c|}{segmentation example: Hi[0] XXX[0] ,[0] its[0] good[0] to[0] hear[0] from[0] you[0] .[1] it[0] sounds[0] like[0]...} \\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Syntactic abstractions of individual words, such as POS tags, have been previously shown to be effective features for similar natural language processing tasks.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger\footnote{\url{https://www.nltk.org/}}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause or phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic lexico-semantic units. Therefore, it is important to represent a word in such a way that preserves all relevant lexical and semantic information. Dense real-valued vector (i.e. embedding) is a form of distributed word representation, in which each word is associated with a vector in a low-dimensional vector space. Embeddings have been previously shown to effectively capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized word embeddings pre-trained on Google News corpus consisting of 1.6 billion words using word2vec software package.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} We skipped the words or punctuation marks, for which pre-trained embeddings could not be found. CRF utilized lexical features and POS tags.

@@

We experimented with 4 different classifiers, including one traditional machine learning model (CRF) and three deep learning methods (MLP, BRNN and CRNN). Since deep learning architectures provide a flexible mechanism for constructing complex models, we take advantage of this flexibility to test different variations of MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Field}: CRF has been widely used in various sequence annotation NLP tasks, such as part-of-speech tagging.\cite{lafferty2001conditional, hirohata2008identifying} Unlike a maximum entropy Markov model, which uses per-state exponential models for conditional probability of the next state given a current state, CRF model directly estimates a distribution of the entire output sequence conditioned on the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ of output sequence $y$, given input sequence $x$:

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags.

\textbf{Multi-Layer Perceptron}: MLP is a neural network, which consists of multiple fully connected layers map an input onto one or several outputs.\cite{rumelhart1986learning} Figure~\ref{fig:mlp} shows a multi-layer perceptron with a single hidden layer. In MLP networks, there is no cycles or loops and information flows only forward, from the input layer through the hidden layer(s) to the output layer. This study utilized MLP with a nonlinear activation function (rectified linear unit) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we utilized dropout in fully connected layers by randomly hiding neurons during training. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models.

@@

\section*{Conclusion}

%Segmentation is the first step of qualitative analysis of unstructured clinical communications.

Segmentation of e-Coaching emails is an integral part of developing and analyzing e-Coaching behavioral interventions. Although several studies have focused on segmentation problem in biomedical context, they are limited to segmenting clinical text in EHR into sections and sentences, with none of them considering segmentation of text into groups of MI behaviors in the context of MI session discourse analysis. By comparing the performance of machine learning methods for the task of segmentation of e-Coaching emails, we found out that CRNN provides the best performance in terms of all performance metrics. Manual segmentation of e-Coaching data is a very resource-intensive and time-consuming task, which can significantly decrease the time and effort required to develop an effective behavioral intervention. Our proposed methods can help to identify textual segments corresponding to MI behaviors in unstructured clinical dialog, which can then be annotated with a classification model in a pipeline setting. Automated segmentation and analysis of e-Coaching emails can significantly decrease the time to identify effective communication strategies in e-mail based motivational interviewing.
